Annotation with protein abundance

```{r, output=FALSE, warning=FALSE}
library(dplyr)
library(Rtsne)
library(patchwork)
library(factoextra)
library(DropletUtils)
library(scater)
library(scran)
library(pheatmap)
library(pheatmap)
library(scuttle)
library(bluster)
library(igraph)

set.seed(42)
datasetName = 'PBMC1' # modify this
isotypeControl = '^Ig' # or '^isotype_control'
datasetFolder = './data/'
rawDataDir = paste(datasetFolder, datasetName, '/raw/10X/', sep='')
outDataDir = paste(datasetFolder, datasetName, '/antibody_annotation/', sep='')
```

Loading data:

```{r}
rawData <- read10xCounts(rawDataDir)
sce <- splitAltExps(rawData, rowData(rawData)$Type)
altExpNames(sce)
```


```{r}
altExp(sce)
```

```{r}
counts(altExp(sce))[,1:10]
```

Distribution of the log10-total ADT count across cells, stratified by the identity of the most abundant marker in each cell:

```{r}
adt.counts <- counts(altExp(sce))
top.marker <- rownames(adt.counts)[max.col(t(adt.counts))]
total.count <- colSums(adt.counts)
boxplot(split(log10(total.count), top.marker), ylab="Log-total ADT count", las=2)
```
Number of features detected (i.e., with non-zero counts) in each cell:

```{r}
adt.detected <- colSums(adt.counts > 0)
hist(adt.detected, col='grey', main="", xlab="Number of detected ADTs")
```

Identification and marking of cells having most of the ADTs counts to 0:

```{r}
controls <- grep(isotypeControl, rownames(altExp(sce)))
qc.stats <- cleanTagCounts(altExp(sce), controls=controls)
summary(qc.stats$zero.ambient) # libraries removed with no ambient contamination
```

```{r}
summary(qc.stats$high.controls)
```

```{r}
hist(log10(qc.stats$sum.controls + 1), col='grey', breaks=50,
    main="", xlab="Log-total count for controls per cell")
thresholds <- attr(qc.stats$high.controls, "thresholds")
abline(v=log10(thresholds["higher"]+1), col="red", lty=2)
```
Deletion of low quality cells (with high percentage of mithocondrial genes):

```{r}
mito <- grep("^MT-", rowData(sce)$Symbol)
df <- perCellQCMetrics(sce, subsets=list(Mito=mito))
mito.discard <- isOutlier(df$subsets_Mito_percent, type="higher")
summary(mito.discard)
```

```{r}
unfiltered <- sce
discard <- qc.stats$discard | mito.discard
sce <- sce[,!discard]
ncol(sce)
```

Computation of library size factors:

```{r}
sf.lib <- librarySizeFactors(altExp(sce))
summary(sf.lib)
```

Estimation of the baseline abundance profile from the ADT count matrix by assuming that the distribution of abundances for each ADT should be bimodal:

```{r}
baseline <- ambientProfileBimodal(altExp(sce))
head(baseline)
```

```{r}
plotExpression(altExp(sce), features=rownames(altExp(sce)), exprs_values="counts") +
    scale_y_log10() + 
    geom_point(data=data.frame(x=names(baseline), y=baseline), mapping=aes(x=x, y=y), cex=3)
```
Computation of each cell size factors as the median of the ratios of that cell’s counts to the baseline profile:

```{r}
sf.amb <- medianSizeFactors(altExp(sce), reference=baseline)
summary(sf.amb)
```

Factors visualization:

```{r}
tagdata <- logNormCounts(altExp(sce))
clusters <- clusterCells(tagdata, assay.type="logcounts") 

by.clust <- split(log2(sf.amb/sf.lib), clusters)
boxplot(by.clust, xlab="Cluster", ylab="Log-ratio (median-based/library size factors)")
```
Computation of alternative size factors from the IgG controls:

```{r}
controls <- grep(isotypeControl, rownames(altExp(sce)))
sf.control <- librarySizeFactors(altExp(sce), subset_row=controls) 
summary(sf.control)
```

```{r}
plot(sf.amb, sf.control, log="xy", 
    xlab="median-based size factors (tag)",
    ylab="Control size factors (tag)")
abline(0, 1, col="grey", lty=2)
```

Setting the size factors for the ADT data:

```{r}
sizeFactors(altExp(sce)) <- sf.amb
```

Performing scaling normalization and log-transformation for both the endogenous transcripts and the ADTs using their respective size factors:

```{r}
sce <- applySCE(sce, logNormCounts)

# Checking that we have normalized values:
assayNames(sce)
```
```{r}
assayNames(altExp(sce))
```

Performing the clustering and visualizing cells in a t-SNE plot:

```{r}
#clusterParams = NNGraphParam(cluster.fun="louvain", cluster.args=list(resolution=2))
#clusterParams = NNGraphParam(cluster.fun=cluster_walktrap, cluster.args=list(steps=2))
clusters.adt <- clusterCells(altExp(sce), assay.type="logcounts") #, BLUSPARAM=clusterParams)
altExp(sce) <- runTSNE(altExp(sce))
colLabels(altExp(sce)) <- factor(clusters.adt)
plotTSNE(altExp(sce), colour_by="label", text_by="label", text_color="red")
```

Sorting antibodies according to their std (from higher to lower):

```{r}
sceCounts = counts(altExp(sce))
std <- apply(sceCounts, 1, sd)
sortedStd = sort(std, decreasing = TRUE)
names(sortedStd)
```
Creation of a heatmap of the average log-abundance of each tag:


```{r}
se.averaged <- sumCountsAcrossCells(altExp(sce), clusters.adt,
    exprs_values="logcounts", average=TRUE)
averaged <- assay(se.averaged)
heatmap_data <- averaged - rowMeans(averaged)

# sort rows according to std
heatmap_data <- heatmap_data[match(names(sortedStd), rownames(heatmap_data)),]

# label heatmap cols with "cluster_id (cluster_size)"
plotLabels = c()
clusters_sizes = table(clusters.adt)
for (i in 1:nrow(clusters_sizes)) {
  plotLabels = c(plotLabels, paste(i, ' (', clusters_sizes[[i]], ')', sep=''))
}

hmap = pheatmap(
  heatmap_data,
  breaks = seq(-3, 3, length.out = 101),
  labels_col = plotLabels,
  cluster_rows = FALSE  # Prevent additional row clustering
)
ggsave(paste(outDataDir, "heatmap.png", sep=""), hmap, height = 8)
```
Saving labels:

```{r}
cellsBarcodes = colData(sce)[,2]
cellsBarcodes <- substr(cellsBarcodes, 1, nchar(cellsBarcodes) - 2)
labels = data.frame("cell" = cellsBarcodes, "cluster.ids" = clusters.adt)
write.csv(labels, paste(outDataDir, "antibody_labels.csv", sep=""), row.names = FALSE)
uclust = sort(unique(clusters.adt))
mapping = data.frame("go" = paste("Cluster-", uclust, sep=""), "id" = uclust)
write.csv(mapping, paste(outDataDir, "antibody_mapping.csv", sep=""), row.names = FALSE)
```

```{r}
sessionInfo()
```